Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2023-03-15 analysis/paper_draft3.Rmd
revised figure captions

---
title: "Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
blinded: 0
authors:
- name: "[Sachin S. Pandya](https://orcid.org/0000-0001-7387-1307)"
  email: sachin.pandya@uconn.edu
- name: "Xiaomeng Li"
  affiliation: University of Connecticut
- name: "[Eric Barón](https://orcid.org/0000-0003-1934-1932)"
  affiliation: University of Connecticut
- name: "[Timothy E. Moore](https://orcid.org/0000-0002-9576-0517)"
  affiliation: University of Connecticut
output:
  bookdown::pdf_book: 
      base_format: rticles::asa_article
      citation_package: "default"
      keep_tex: TRUE
header-includes: 
  - \usepackage{amsmath}
  - \usepackage{subfig}
#fontsize: 12pt
bibliography: references.bib
biblio-style: agsm.bst
nocite: |
  @WashR37, @calccp2021
keywords: "Batson challenge, peremptory strikes, power prior, Bayesian"
abstract: |
  US law bars using peremptory strikes during jury selection because of prospective juror race, ethnicity, sex, or membership in certain other cognizable classes. Here, we extend a Bayesian approach for detecting such illegal strike bias by showing how to incorporate historical data on an attorney's use of peremptory strikes in past cases. In so doing, we use the power prior to adjust the weight of such historical information in the analysis. Using simulations, we show how the choice of the power prior's discounting parameter influences bias detection (how likely the credible interval for the bias parameter excludes zero), depending on the degree of incompatibility between current and historical trial data. Finally, we extend this approach with a prototype software application that lawyers could use to detect strike bias in real time during jury-selection. We illustrate this application's use with real historical strike data from a convenience sample of cases from one court.

@@

```{r figbd9515, out.width = "95%", fig.align='center', fig.cap = "Bias detection rate for simulations based on 15 strikes. Rows of larger squares correspond to combinations of alpha values (0.1, 0.5, and 1) for low, median and equal weight of historical data to current trial data. Columns of larger squares correspond to 1, 2, or 3 historical trials. Small squares correspond to combinations of bias in current (rows) and historical (columns) trials."}

@@

```{r figbd9510, out.width = "95%", fig.cap = "Bias Detection Rates (95 pct credible interval, 10 strikes scenario).", fig.align='center'}

# Bias detection rate for simulations based on 10 strikes. Rows of larger squares correspond to combinations of alpha valuesa (0.1, 0.5, and 1) for low, median and equal weight of historical data to current trial data. Columns of larger squares correspond to 1, 2, or 3 historical trials. Small squares correspond to combinations of bias in current (rows) and historical (columns) trials.

@@

```{r figbd956, out.width = "95%", fig.cap = "Bias Detection Rates based on 95 pct credible interval for 6 strike scenario.", fig.align='center'}

@@

```{r figsdprior, out.width = "80%", fig.cap = "Sensitivity to Standard Deviation for Initial Prior by alpha.", fig.align='center'}

@@

```{r figcmptbly, out.width = "85%", fig.cap = "Bias detection under compatibility scenarios (equal bias in current and historical trials) for different numbers of strikes (columns), sizes of historical data (rows) and levels of alpha (shapes).", fig.align='center'}

>>>>

---
title: "Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
blinded: 0
authors:
- name: "[Sachin S. Pandya](https://orcid.org/0000-0001-7387-1307)"
  affiliation: University of Connecticut
  email: sachin.pandya@uconn.edu
- name: "Xiaomeng Li"
  affiliation: University of Connecticut
- name: "[Eric Barón](https://orcid.org/0000-0003-1934-1932)"
  affiliation: University of Connecticut
- name: "[Timothy E. Moore](https://orcid.org/0000-0002-9576-0517)"
  affiliation: University of Connecticut
output:
  bookdown::pdf_book: 
      base_format: rticles::asa_article
      citation_package: "default"
      keep_tex: TRUE
header-includes: 
  - \usepackage{amsmath}
  - \usepackage{subfig}
#fontsize: 12pt
bibliography: references.bib
biblio-style: agsm.bst
nocite: |
  @WashR37, @calccp2021
keywords: "Batson challenge, peremptory strikes, power prior, Bayesian"
abstract: |
  US law bars using peremptory strikes during jury selection because of prospective juror race, ethnicity, sex, or membership in certain other cognizable classes. Here, we extend a Bayesian approach for detecting such illegal strike bias by showing how to incorporate historical data on an attorney's use of peremptory strikes in past cases. In so doing, we use the power prior to adjust the weight of such historical information in the analysis. Using simulations, we show how the choice of the power prior's discounting parameter influences bias detection (how likely the credible interval for the bias parameter excludes zero), depending on the degree of incompatibility between current and historical trial data. Finally, we extend this approach with a prototype software application that lawyers could use to detect strike bias in real time during jury-selection. We illustrate this application's use with real historical strike data from a convenience sample of cases from one court.

@@

```{r figbd9515, out.width = "95%", fig.align='center', fig.cap = "Bias detection rates (proportion of simulations where 0 is excluded from the 95 percent credible interval of bias parameter) for simulations based on 15 strikes. Rows of larger squares correspond to combinations of alpha values (0.1, 0.5, and 1) for low, median and equal weight of historical data to current trial data. Columns of larger squares correspond to 1, 2, or 3 historical trials. Small squares correspond to combinations of bias in current (rows) and historical (columns) trials."}

@@

```{r figbd9510, out.width = "95%", fig.cap = "Bias detection rates (proportion of simulations where 0 is excluded from the 95 percent credible interval of bias parameter) for simulations based on 10 strikes.", fig.align='center'}

@@

```{r figbd956, out.width = "95%", fig.cap = "Bias detection rates (proportion of simulations where 0 is excluded from the 95 percent credible interval of bias parameter) for simulations based on 6 strikes.", fig.align='center'}

@@

```{r figsdprior, out.width = "80%", fig.cap = "Sensitivity to Initial Prior's Standard Deviation (SD) by level of discounting (alpha). Simulations were run with 3 levels of the initial prior SD: low (SD = 1), medium (SD = 2), and hight (SD = 100). Bias detection rate corresponds to the proportion of simulations in which bais was detected based on 0 being excluded from the 95 percent credible interval. Bias was equal in current and historical trials and three levels were selected: 0 (no bias scenario), 1, and 2. In general, bias detection is relatively unaffected by choice initial SD, especially when alpha is close to 1.", fig.align='center'}

@@

```{r figcmptbly, out.width = "85%", fig.cap = "Bias detection under compatibility scenarios (equal bias in current and historical trials) for different numbers of strikes (columns), sizes of historical data (rows) and levels of alpha (shapes). As the weight applied to historical data increases (higher alpha), bias detection increases. Bias detection rates also increase as the number of strikes and historical datasets increase.", fig.align='center'}
